Evolutionary rates of SARS-CoV-2

### 2022-07-29 manuscript/variant_rates.tex
add bib etc

\documentclass[aps,rmp, onecolumn]{revtex4}
%\documentclass[a4paper,10pt]{scrartcl}
%\documentclass[aps,rmp,twocolumn]{revtex4}

\usepackage[utf8]{inputenc}
\usepackage{amsmath,graphicx}
\usepackage{color}
%\usepackage{cite}

@@

\bibliography{bib}

@@

Since its emergence in late 2019, SARS-CoV-2 has displayed a discontinuous pattern of evolution with large unobserved jumps in sequence space giving rise to new variants \citep{alpha,beta,gamma,delta,omicron}.
\begin{itemize}
    \item hypothesis of the jumps
    \item link to other examples of decaying rate of evolution H1N1pdm, HIV within host
    \item
\end{itemize}

@@

Due to shared ancestry, a divergence (or root-to-tip) regression against time is not a suitable method to determine the evolutionary rate.
However, in the case of rapidly expanding variants we typically observe a large number of independent lineages emanating from a basal polytomy.
Along each of these lineages, mutation accumulation is independent.
\begin{figure*}
    \includegraphics[width=\textwidth]{figures/20I_rtt.png}
    \caption{{\bf Divergence increases linearly with time (20I (Alpha)).} Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
    The blue line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter. Each panel also shows mean $\pm$ standard deviation and a weighted linear fit.}
\end{figure*}

\end{document}

>>>>

\documentclass[aps,rmp, twocolumn]{revtex4}
%\documentclass[a4paper,10pt]{scrartcl}
%\documentclass[aps,rmp,twocolumn]{revtex4}

\usepackage[utf8]{inputenc}
\usepackage{amsmath,graphicx}
\usepackage{color}
\usepackage{booktabs}
%\usepackage{cite}

@@

@@

Since its emergence in late 2019, SARS-CoV-2 has displayed a discontinuous pattern of evolution with large unobserved jumps in sequence space giving rise to phylogenetically distinct variants \citep{hodcroft_spread_2021,volz_assessing_2021,tegally_detection_2021,faria_genomics_2021,naveca_covid-19_2021,viana_rapid_2022}.
Many of these variants spread considerably faster and quickly displaced the resident variants either because of intrinsically increased transmissibility, evasion of prior immunity in the population, or a combination of both.
Variants of Concern or Interest were designated by the WHO and labeled by Greek letters \citep{konings_sars-cov-2_2021}.
The branches leading to these variants are characterized by many amino acid changing mutations that often cluster in the S1 domain of the spike protein \citep{kistler_rapid_2022}.
This pattern of rapid non-synonymous evolution in viral surface proteins that interact with the host cells in common among many RNA viruses and for example well studied in influenza virus A evolution \citep{bhatt_genomic_2011,strelkowa_clonal_2012}.

In contrast to influenza virus evolution, the evolution of SARS-CoV-2 has been characterized by unusually large jumps in sequence space.
New variants with 10s of previously unobserved mutations emerged suddenly without intermediate genomes being observed, most dramatically with the emergence of Omicron in later 2021 \citep{viana_rapid_2022}.
The pre-dominant hypothesis for this cryptic emergence of highly mutated, transmissible and immune evasive variants are chronic infections that are frequently observed in patients with impaired immune systems, either through HIV-1 infection \citep{cele_sars-cov-2_2022} or medical intervention \citep{choi_persistence_2020,kemp_sars-cov-2_2021}.
Onward transmission from such chronic infections has also been documented \citep{gonzalez-reiche_intrahost_2022}.
However, to date there is no direct evidence for the emergence of any variant and the degree to which chronic infections might have contributed is a matter of debate.
The case for chronic infection being an important contributor is strongest for the variants Alpha and Omicron
\citep{hill_origins_2022}.

Here, we investigate the rate at which the SARS-CoV-2 genome changes within variants and compare this rate to the rate at which the variants themselves accumulate differences relative to reference sequence.

\section*{Results}

@@

Due to shared ancestry, a divergence (or root-to-tip) regression against time is not a suitable method to determine the evolutionary rate.
However, in the case of rapidly expanding variants we typically observe a large number of independent lineages emanating from a basal polytomy.
Along each of these lineages, mutation accumulation is independent and divergence increase allows to estimate the rate.
Fig.~\ref{fig:rate_alpha} shows the increasing intra-clade divergence for clade 20I corresponding to VoC Alpha.

\begin{figure*}[tb]
    \includegraphics[width=\textwidth]{figures/20I_rtt.png}
    \caption{{\bf Divergence increases linearly with time (20I (Alpha)).} Each panel shows the number of within-clade mutations (total (A), amino acid changing (B), synonymous (C)) as a function of time.
    The blue line in panel A indicates the divergence cut-off, panels B\&C only show sequences that pass the divergence filter. Each panel also shows mean $\pm$ standard deviation and a weighted linear fit.
    \label{fig:rate_alpha}}
\end{figure*}

For almost all major Nextstrain clades, the average divergence increases linearly in time (similar patterns as Fig.~\ref{fig:rate_alpha}) allowing us to estimate clade specific evolutionary rates for amino acid and silent changes.
These rates are summarized in Fig.~\ref{fig:rate_summary}.

\begin{figure*}
    \includegraphics[width=\textwidth]{figures/rate_summary.png}
    \caption[]{{\bf Divergence and evolutionary rates of different Nextstrain clades.} Panels A,B,\&C show the estimated divergence of the founder genotype of each clade (big dot) and the subsequent divergence trend for all nucleotide changes, amino acid changes, and synonymous changes, respectively. In addition, each panel contains a regression of the divergence of clade founders vs time (gray line).
    The standard deviation expected based on Poisson statistics is indicated as shared area.
    Panel D summarizes the individual rate estimate (dots) and compares them to the estimate inter-clade rates (gray lines).
    \label{fig:rate_summary} }
\end{figure*}

Rates of synonymous change are very consistent across variants and also agree with the overall rate of synonymous changes of about 5-8 changes per genome per year.
Rates of non-synonymous changes are much more variable.
Within clades, the rate of non-synonymous changes varies between 5 and 16 changes per year.
Earlier clades are estimated to have larger rates around 10-15 changes per year, while rate estimates for later clade fall  between 3 to 9 changes per year.
In contrast, the int-clade non-synonymous rate exceeds 25 changes per year.

\begin{table*}
\begin{tabular}{l|rrrrrr}
    \hline
    {\bf clade} &  overall rate $[y^{-1}]$ & aa rate $[y^{-1}]$ &  syn rate $[y^{-1}]$ & overall div. &  aa div. &  syn div. \\
    \hline
      19A &     19.21 &   11.85 &     7.36 &     0 &       0 &        0 \\
      19B &     24.40 &   14.50 &     9.90 &     2 &       1 &        1 \\
      20A &     20.79 &   11.44 &     9.35 &     4 &       2 &        2 \\
      20B &     14.64 &    8.42 &     6.22 &     7 &       4 &        3 \\
      20C &     19.22 &   11.64 &     7.57 &     6 &       4 &        2 \\
     20A+ &     18.48 &   10.26 &     8.22 &     4 &       2 &        2 \\
      20E &     13.77 &    6.04 &     7.73 &    11 &       4 &        7 \\
      20H &     11.16 &    5.22 &     5.94 &    19 &      15 &        4 \\
      20I &     11.65 &    5.77 &     5.88 &    28 &      18 &       10 \\
      20J &     10.93 &    2.66 &     8.28 &    33 &      21 &       12 \\
      21I &     16.53 &    8.88 &     7.65 &    28 &      23 &        5 \\
      21J &     16.20 &    8.19 &     8.00 &    32 &      27 &        5 \\
      21K &     12.74 &    5.41 &     7.34 &    54 &      45 &        9 \\
      21L &      8.66 &    4.67 &     3.99 &    66 &      49 &       17 \\
      22A &      8.46 &    2.92 &     5.55 &    69 &      51 &       18 \\
      22B &     13.78 &    4.13 &     9.65 &    64 &      49 &       15 \\
      \hline
    \end{tabular}
\caption{{\bf Evolutionary rates estimates from root-to-tip regressions for overall nucleotide changes, amino acid changes, and synonymous changes.} \label{tab:rates}}
\end{table*}

\begin{figure}
    \includegraphics[width=0.5\textwidth]{figures/rate_progression.pdf}
    \caption[]{{\bf Divergence and evolutionary rates of different Nextstrain clades over time.} Synonymous rates estimates are stable in time and fluctuate around the rate estimates for between clades. Non-synonymous rate estimates are highest for clades 19A - 20C.
    \label{fig:rate_progression} }
\end{figure}

\subsection*{Diversification within variants}

\bibliography{bib}
\end{document}
